---
title: "Verification of Hypothesis About How the Change of Dosage and Type Supplement for Vitamin C Affects the Tooth Growth in Guinea Pigs"
author: Daniele Francesconi
output: html_document
date: "November 16th, 2016"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
In this document is reported the second part of the peer-graded assignment for the [Statistical Inference Course](https://www.coursera.org/learn/statistical-inference). In this part of the project the data relative to "The Effect of Vitamin C on Tooth Growth in Guinea Pigs" included in the R dataset `ToothGrowth` will be analysed to see if its two main parameter, dosage and type supplement affect the length of the tooth growth in Guinea Pigs.

## Exploratory Data Analyses

First the data description included in the dataset package:

```
The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).
```

Below the relative data summary:

```{r}
data(ToothGrowth)
head(ToothGrowth)
```

The 3 features of the files are:

* `len` = Tooth length
* `supp` = Supplement type
* `dose` = Dose in milligrams/day

```{r}
summary(ToothGrowth)
str(ToothGrowth)
unique(ToothGrowth$dose)
unique(ToothGrowth$supp)
```

As we can see the `dose` has only three values (0.5, 1.0, 2.0) and the type supplement `supp` two: OJ and VC.
A quick display of the data 

```{r}
require(graphics)
coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth,
       xlab = "ToothGrowth data: length vs dose, given type of supplement")
```

give us the impression that a higher tooth growth is correlated with the increasing dosage. This effect looks more visible with the VC type supplement. It also suggests that type supplements could affect the tooth growth.

## Hypothesis Tests

These are the hypothesis that I am going to verify, including NULL and alternative hypothesis for any respective case:

1 **Type supplement**

 * *NULL hypothesis* > There is no length difference when VC and OJ are used. 
 * *Alternative hypothesis* > More tooth growth increasing with OJ respect to VC.

2 **Dosage** 

 * *NULL hypothesis* > There is no length difference when difference dosage is used by supplement type.
 * *Alternative hypothesis* > More tooth growth increasing for higher dosage (2.0 > 1.0 > 0.5) by supplement type.
 
Before to proceed with the verification of the hypothesis, I formulate some basic assumptions around the data:

- The tooth growth follows a normal distribution
- The variables are independent and identically distributed (iid).
- The relative variance changes against the dosage and supplement type.

I am going to use the R function `t.test(...)` and the p-value returned from it, which is the main index to accept or reject the hypothesis. I will keep as default its parameters for `paired = FALSE`, `var.equal = FALSE` and `conf.level = 0.95`.

### Hypothesis Tests for Type Supplement VC / OJ

To verify the hypothesis relative to type supplement, I start separating in two subsets the data, based on type supplement:

```{r}
tg_oj <- ToothGrowth[ToothGrowth$supp == "OJ",]$len
tg_vc <- ToothGrowth[ToothGrowth$supp == "VC",]$len
```

and I compare the two subsets as independent to get the `p-value`:

```{r}
t.test (tg_oj, tg_vc, alternative = "greater")$p.value
```

The returned p-value is smaller than 0.05. It suggests that I can reject the NULL hypothesis as it is below the default alpha-error (5% tollerance). To conclude I can state that the growth is different based on different type supplement and that the supplement OJ affects the tooth growth more than VC.

### Dosage by Type Supplement

I start to separate data by different dosage for the supplement type OJ:

```{r}
d_05_OJ <- ToothGrowth[ToothGrowth$dose == 0.5 & ToothGrowth$supp == "OJ",]$len
d_10_OJ <- ToothGrowth[ToothGrowth$dose == 1.0 & ToothGrowth$supp == "OJ",]$len
d_20_OJ <- ToothGrowth[ToothGrowth$dose == 2.0 & ToothGrowth$supp == "OJ",]$len
```

In a similar way to it was done before, let's see what the `t.test` returns:

```{r}
t.test (d_05_OJ, d_10_OJ, alternative = "less")$p.value
t.test (d_10_OJ, d_20_OJ, alternative = "less")$p.value
```

As we can see, the p-values (relatively to the alpha-error) are still below the confidence level (default = 0.95) 0.05. So once more we can reject the NULL hypothesis and say that for the type supplement OJ there is an increased tooth growth for higher dosages.

For the type supplement VC:

```{r}
d_05_VC <- ToothGrowth[ToothGrowth$dose == 0.5 & ToothGrowth$supp == "VC",]$len
d_10_VC <- ToothGrowth[ToothGrowth$dose == 1.0 & ToothGrowth$supp == "VC",]$len
d_20_VC <- ToothGrowth[ToothGrowth$dose == 2.0 & ToothGrowth$supp == "VC",]$len
```

Similar to the OJ case:

```{r}
t.test (d_05_VC, d_10_VC, alternative = "less")$p.value
t.test (d_10_VC, d_20_VC, alternative = "less")$p.value
```

Based on the p-values, again we can reject the NULL hypothesis and confirm that the increasing of dosage affects the tooth growth for the type supplement VC.