---
title: "Verify that the Change of Dosage and Type Supplement for Vitamin C Affects the Tooth Growth in Guinea Pigs"
author: Daniele Francesconi
output: html_document
date: "November 16th, 2016"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
In this document is reported the second part of the peer-graded assignment for the [Statistical Inference Course](https://www.coursera.org/learn/statistical-inference). In this part of the project the data relative to "The Effect of Vitamin C on Tooth Growth in Guinea Pigs" included in the R dataset `ToothGrowth`.

## Exploratory Data Analyses

First the data description included in the dataset package:
```
The response is the length of odontoblasts (cells responsible for tooth growth) in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) by one of two delivery methods, (orange juice or ascorbic acid (a form of vitamin C and coded as VC).
```

```{r}
data(ToothGrowth)
head(ToothGrowth)
```

The 3 features of the files are:

* len = Tooth length
* supp = Supplement type
* dose = Dose in milligrams/day

```{r}
summary(ToothGrowth)
str(ToothGrowth)
unique(ToothGrowth$dose)
unique(ToothGrowth$supp)
```

As we can see the `dose` has only three values (0.5, 1.0, 2.0) and the supplement two: OJ and VC.
A quick data representation

```{r}
require(graphics)
coplot(len ~ dose | supp, data = ToothGrowth, panel = panel.smooth,
       xlab = "ToothGrowth data: length vs dose, given type of supplement")
```

give us the impression that the tooth growth is correlated with the dose: higher dose, more growth. This effect looks more visible with the VC type supplement.

## Hypothesis Tests

I will verify the following hypothesis defining what are my NULL hypothesis for any respective case:

* **Type supplement** There is no length difference when VC and OJ are used. The alternate hypotesis says that the length with OJ > VC.
* **Dosage** There is no length difference when difference dosage is used by supplement type. The alternate hypotesis says that the length with dosage at 2.0 > 1.0 > 0.5.

The basic assumptions are that the variables are independent and identically distributed (iid). The tooth growth follows a normal distribution and the relative variance changes against the dosage and supplement type.

### Hypothesis Tests for Type Supplement VC / OJ

To verify the hypothesis I used the R function `t.test(...)`. First I separated two subsets based on type supplement:

```{r}
tg_oj <- ToothGrowth[ToothGrowth$supp == "OJ",]$len
tg_vc <- ToothGrowth[ToothGrowth$supp == "VC",]$len
```

and I compared the two subsets as independent (default condition for the `t.test` function) to get the `p-value`:

```{r}
t.test (tg_oj, tg_vc, alternative = "greater")$p.value
```

The returned p-value is smaller than 0.05. It suggests that I can reject the NULL hypothesis as it is below the default alpha-error (5% tollerance). To conclude I can state that the growth is different based on different type supplement and that the supplement OJ affects the tooth growth more than VC.

### Dosage by Type Supplement

I start to isolate data by different dosage for the supplement type OJ:

```{r}
d_05_OJ <- ToothGrowth[ToothGrowth$dose == 0.5 & ToothGrowth$supp == "OJ",]$len
d_10_OJ <- ToothGrowth[ToothGrowth$dose == 1.0 & ToothGrowth$supp == "OJ",]$len
d_20_OJ <- ToothGrowth[ToothGrowth$dose == 2.0 & ToothGrowth$supp == "OJ",]$len
```

How it has been done before, let's see what the `t.test` returns:

```{r}
t.test (d_05_OJ, d_10_OJ, alternative = "less")$p.value
t.test (d_10_OJ, d_20_OJ, alternative = "less")$p.value
```

As we can see, the p-values (relatively to the alpha-error) are still below the confidence level (default = 0.95). So once more we can reject the NULL hypothesis and say that for the type supplement OJ there is an increased tooth growth for higher dosages.

For the type supplement VC:

```{r}
d_05_VC <- ToothGrowth[ToothGrowth$dose == 0.5 & ToothGrowth$supp == "VC",]$len
d_10_VC <- ToothGrowth[ToothGrowth$dose == 1.0 & ToothGrowth$supp == "VC",]$len
d_20_VC <- ToothGrowth[ToothGrowth$dose == 2.0 & ToothGrowth$supp == "VC",]$len
```

Similar to the OJ case:

```{r}
t.test (d_05_VC, d_10_VC, alternative = "less")$p.value
t.test (d_10_VC, d_20_VC, alternative = "less")$p.value
```

Based on the p-values, again we can reject the NULL hypothesis and confirm that the increasing dosage affects the tooth growth for the type supplement VC.