---
title: "Regression Models Course Project"
author: "Daniele Francesconi"
date: "23/12/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

TBD

## Data Summary

The data included in the R's datasets is called `mtcars` and the description says:

> The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973–74 models).

Before a quick look into the data, dimensions and how the data look like (type included):

```{r mtcars}
data("mtcars")
dim(mtcars)
head(mtcars)
summary(mtcars)
```

So we have 32 observations on 11 variables. From the documentation I have extracted the following variables descriptions:

Variables|Description
---------|-----------
*mpg*	| Miles/(US) gallon
*cyl*	| Number of cylinders
*disp* | Displacement (cu.in.)
*hp*	| Gross horsepower
*drat*	| Rear axle ratio
*wt*	| Weight (1000 lbs)
*qsec*	| 1/4 mile time
*vs*	| V/S
*am*	| Transmission (0 = automatic, 1 = manual)
*gear*	| Number of forward gears
*carb*	| Number of carburetors

Some of the values are *discrete* like: `cyl`, `vs`, `am`, `gear` and `carb`. However I will consider only `vs` and `am` as `factor`s:

```{r}
mtcars$vs <- factor(mtcars$vs,labels=c('V','S'))
mtcars$am <- factor(mtcars$am,labels=c('A','M'))
```

## Exploratory Data Analysis

Let's see how `cyl`, `vs`, `gear` and `carb` are distributed against `am` (that is the parameter that represents our main interest along `mpg`):

```{r, out.width = '100%', dpi=200, cache=TRUE}
par(mfrow=c(2,2))

display_barplot <- function(table, title, xlab) {
  barplot(table, main=title, xlab=xlab, col=c("darkblue","red"), beside=TRUE)
  legend("topright", title = "AM", rownames(table), fill = 1:2, ncol = 1, cex = 0.5)
}

display_barplot(table(mtcars$am, mtcars$cyl) ,"AM vs. Cylinders", "Number of Cylinders")
display_barplot(table(mtcars$am, mtcars$vs) ,"AM vs. VS", "V/Straight Engine")
display_barplot(table(mtcars$am, mtcars$gear) ,"AM vs. Gears", "Number of Gears")
display_barplot(table(mtcars$am, mtcars$carb) ,"AM vs. Carb", "Number of Carburetors")
```

Gears looks not having the fifth one in the Automatic model. Similarly Carburetors looks limited in the case of automatic model.

Now, let's have a summary of the data, included two scatterplot matrixes relative to the variables for a better data insight: `mpg`, `hp` and `wt` separated by `am` (tranmission) and `cyl`:

```{r, out.width = '100%', dpi=200}
library(car)
scatterplotMatrix(~mpg+hp+wt|am, data=mtcars, main="MPG - HP - WT vs AM")
```

On the first quadrant we can see that we have higher values of `mpg` for manual transmission. This trend looks is confirmed for `mpg` against `hp` and `wt`.

```{r, out.width = '100%', dpi=200}
scatterplotMatrix(~mpg+hp+wt|cyl, data=mtcars, main="MPG - HP - WT vs CYL")
```

Here we have 3 main categories of cars based on the number cylinders: 4, 6 and 8. It is easy to see that for higher number of cylinders `mpg` drops down due to higher fuel consumption. Same observation can be considered for higher values of `hp` and `wt` that correspond to lower `mpg` values.

In **Appendix A** there is a complete display of the correlations across the whole set of variables in the dataset. Looking at the values for `mpg`, some of those correlations looks negative. However these correlations are meant for single independent variables. In case of multi-regressors, their impact on the outcome might change. Indeed, correlations suggest us another main point: `mpg` is strongly correlated to `am` but also `cyl`, `disp`, `hp`, `drat`, `wt` and `vs`. What makes the things even more complex are the mutual interactions between these variables. Using the *Variance Inflation Factors*, we can see that some variables are more influenced by others:

```{r}
sort(sqrt(vif(lm(mpg~., mtcars))), decreasing = TRUE)
sort(sqrt(vif(lm(mpg~. -wt, mtcars))), decreasing = TRUE)
```

`disp` has the highest value, but it drops down removing `wt`. It suggests a interaction between the two.

## Model Selection

To select the model, we will consider the fix regressor `am` for the outcome `mpg` as the base model. I will add incrementally new parameters, based to their absolute correlation respect to theh graph in **Appendix A** creating a set of nested models. Using the `anova` function I will select the best model, related to the `p-value`:

```{r}
fit1  <- lm(mpg ~ am , mtcars)
fit2  <- update(fit1, mpg ~ am + wt)
fit3  <- update(fit1, mpg ~ am + wt + cyl)
fit4  <- update(fit1, mpg ~ am + wt + cyl + disp)
fit5  <- update(fit1, mpg ~ am + wt + cyl + disp + hp)
fit6  <- update(fit1, mpg ~ am + wt + cyl + disp + hp + drat)
fit7  <- update(fit1, mpg ~ am + wt + cyl + disp + hp + drat + carb)
fit8  <- update(fit1, mpg ~ am + wt + cyl + disp + hp + drat + carb + gear)
fit9  <- update(fit1, mpg ~ am + wt + cyl + disp + hp + drat + carb + gear + qsec)
fit10 <- update(fit1, mpg ~ am + wt + cyl + disp + hp + drat + carb + gear + qsec + vs)

anova(fit1, fit2, fit3, fit4, fit5, fit6, fit7, fit8, fit9, fit10)
```

The model based on `am + wt` looks the best one. Still good is the model `am + wt + cyl`. Let us see if it can be improved including the interaction `wt * cyl`:

```{r}
fit3Interaction  <- update(fit1, mpg ~ am + wt + cyl + wt * cyl)

anova(fit1, fit2, fit3, fit3Interaction)
```

There is not a real vantage with the interaction. So I will stick with the model `mpg ~ am + wt` only.

## Inference

Comparing the coefficients for the model `mpg ~ am` and `mpg ~ am + wt`:

```{r}
summary(fit1)$coefficients
summary(fit2)$coefficients
```

It is possible to see that in the first case the manual transmission is better respect to the automatic one, as the coefficient increase is +7.245. In the second case, the two transmissions are pretty much the same. It means that the a `wt` regressor affects a lot the transmission.

It means that we cannot easily stating if one transmission is better than another.

## Residual Analysis

In **Appendix B** residual plots for the models `fit2` and `fit10` are displayed. In both cases it seems that both models fit enough well the data. However there are a few values that are highlighted. It can be seen clearly from the graph *Normal Q-Q* as those value are out from the predicted linear model (suspiciously have some leverage). Instead the *Residuals-Fitted* graph looks covering well the distribution of the data. 

The suspicious values are related to the cars: Chrysler Imperial, Toyota Corolla and Fiat 128. Let's see their influence using *hatvalues* to see their level of leverage and *dfbetas* too see how much their inclusion affects the slope coefficient:

```{r, out.width = '100%', out.height = '100%', dpi=200, cache = TRUE, echo = FALSE}
require(reshape2)

model_under_investigation <- lm (mpg ~ am + wt - 1, mtcars)

htv <- melt(as.matrix(round(hatvalues(model_under_investigation), 3)))
ggplot(data=htv, aes(x=Var1, y=value, group=Var2)) + geom_line() + theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5)) + ylab('hatvalues') + xlab('cars') + ggtitle("Hatvalues: mpg ~ am + wt")

dfb <- melt(round(dfbetas(model_under_investigation), 3))
ggplot(data=dfb, aes(x=Var1, y=value, group=Var2)) + geom_line(aes(colour = Var2))+ theme(axis.text.x = element_text(angle=90, hjust=1, vjust=0.5)) + ylab('dfbetas') + xlab('cars') + ggtitle("Dfbetas: mpg ~ am + wt")
```

For the *hatvalues* values are not small, but still in the average.
In the case of *dbetas* values, the 3 cars looks to have high peaks, and in the case of Chrysler Imperial, both `am` types, the peaks are negative against the positive for `wt`. It would explain why it is in the top left quadrant.
We can conclude that those values have some influence but they do not affect particularly the model.

It is also interesting to see the graphical comparison of the models `mpg ~ am`, `mpg ~ am + wt` and `mpg ~ .`:

```{r}
plot(predict(fit2))
lines(predict(fit2), col = "black")
lines(predict(fit1), col = "blue")
lines(predict(lm(mpg ~ ., mtcars)), col = "red")
```

Overall we might to say that our model `fit2` it is more closed to the `fit10`, indicating that `am` and `wt` are quite important factors.

## Conclusions

## Appendix A

```{r, out.width = '100%', out.height = '100%', dpi=200, cache = TRUE, echo = FALSE}
require(GGally)
require(ggplot2)

lowerFn <- function(data, mapping, ...) {
  p <- ggplot(data = data, mapping = mapping) +
    geom_point(color = 'blue', alpha=0.3, size=0.5) +
    geom_smooth(color = 'black', method='lm', size=0.7,...)
  p
}

printCorrelations <- function(modelData, graphTitle) {
  g <- ggpairs( 
    data = modelData,
    title = graphTitle,
    lower = list(
      continuous = wrap(lowerFn) #wrap("smooth", alpha = 0.3, color = "blue", lwd=1) 
    ),
    upper = list(continuous = wrap("cor", size = 2))
  )
  g <- g + theme(
    axis.text = element_text(size = 4),
    axis.title = element_text(size = 4),
    legend.background = element_rect(fill = "white"),
    panel.grid.major = element_line(colour = NA),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "grey95")
  )
  suppressMessages(print(g)) #, bottomHeightProportion = 0.5, leftWidthProportion = .5))
}

printCorrelations( subset(mtcars, select = c(mpg,am)), "Correlation Matrix for MPG/AM" )
printCorrelations( mtcars, "Correlation Martrix for all variables" )
```

## Appendix B

### Model mpg ~ am + wt

```{r, out.width = '100%', out.height = '100%', dpi=200, cache = TRUE, echo = FALSE}
plot(fit2)
```

### Model mpg ~ . (in 4 quadrants)

```{r, out.width = '100%', out.height = '100%', dpi=200, cache = TRUE, echo = FALSE}
par(mfrow=c(2,2))
plot(fit10)
```